---
title: "Project 1"
author: "George James"
format: 
  html:
    css: styles.css  
    toc: true
    toc-location: left
---

This is my first mini-project for Data Science 2. I looked at coffee prices and how they differed by state.

```{r}
#| message: false
#| warning: false

# Load initial packages required
library(tidyverse)
library(mdsr) 
#coffee_data:here("https://vinepair.com/articles/price-of-coffee-state-map/") 
# Create a data frame with coffee prices by state
coffee_data <- data.frame(
  State = c("Hawaii", "California", "Washington", "Arizona", "Massachusetts",
            "Colorado", "Utah", "Florida", "Vermont", "Connecticut",
            "New Jersey", "New Mexico", "Alaska", "Louisiana", "Maryland",
            "New York", "Nevada", "Oregon", "South Carolina", "New Hampshire",
            "Rhode Island", "Maine", "Virginia", "Illinois", "Kentucky",
            "Georgia", "Texas", "Delaware", "Idaho", "North Carolina",
            "Pennsylvania", "Tennessee", "Oklahoma", "Minnesota", "Wyoming",
            "Missouri", "Alabama", "Arkansas", "Michigan", "Ohio",
            "Indiana", "Mississippi", "Wisconsin", "Iowa", "South Dakota",
            "West Virginia", "North Dakota", "Kansas", "Montana", "Nebraska"),
  
  Price_of_Coffee = c(4.98, 3.88, 3.69, 3.51, 3.49, 3.43, 3.40, 3.38,
                      3.32, 3.31, 3.31, 3.31, 3.27, 3.26, 3.25, 3.24,
                      3.23, 3.21, 3.17, 3.13, 3.13, 3.08, 3.08, 3.07,
                      3.03, 3.02, 2.99, 2.97, 2.94, 2.94, 2.94, 2.91,
                      2.90, 2.89, 2.89, 2.88, 2.86, 2.86, 2.83, 2.79,
                      2.78, 2.77, 2.76, 2.70, 2.66, 2.65, 2.64, 2.59,
                      2.56, 2.12)
)

# Load US states map data
library(maps)
us_states <- map_data("state")
head(us_states)

# Calculate statistics for coffee prices
mean_price <- mean(coffee_data$Price_of_Coffee)
max_price <- max(coffee_data$Price_of_Coffee)
min_price <- min(coffee_data$Price_of_Coffee)

# Separate coffee data into above and below mean
above_mean <- coffee_data[coffee_data$Price_of_Coffee > mean_price, ]
below_mean <- coffee_data[coffee_data$Price_of_Coffee < mean_price, ]

# Categorize coffee prices as Above or Below Mean
coffee_data <- coffee_data %>%
  mutate(Price_Group = ifelse(Price_of_Coffee > mean_price, "Above Mean", "Below Mean"))

# Merge coffee data with state map data
merged_data <- us_states %>%
  left_join(coffee_data, by = c("region" = "State"))

# Basic US states map
us_states |>
  ggplot(mapping = aes(x = long, y = lat, group = group)) + 
  geom_polygon(fill = "white", color = "black") +
  theme_minimal() +
  labs(title = "US States Map")

# Prepare coffee data for plotting
coffee_data <- coffee_data |>
  mutate(State = str_to_lower(State))

# Create a map of coffee prices by state
coffee_data |>
  right_join(us_states, by = c("State" = "region")) |>
  rename(region = State) |>
  ggplot(mapping = aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = Price_of_Coffee), color = "black") +
  theme_minimal() +
  labs(title = "Coffee Prices by State", fill = "Price of Coffee ($)")

# Merge coffee data with state map data
merged_data0 <- us_states %>%
  left_join(coffee_data, by = c("region" = "State"))
```
A map of the United States displaying coffee prices by state. Each state is filled with a gradient color representing the price of coffee, with darker colors indicating higher prices. A color legend on the side shows the corresponding price ranges. We can see that Nebraska is the darkest, meaning it has the lowest coffee price. California is the lightest along with Washinton, indicating that they have higher coffee prices. 

# Final map showing coffee price groups
```{r}
ggplot(data = merged_data0, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = Price_Group), color = "darkgrey", linewidth = 0.2) +
  labs(fill = "Coffee Price Group") +
  coord_fixed() + 
  theme_void() +  
  scale_fill_manual(values = c("brown1", "brown4")) +
  ggtitle("Coffee Prices Across the USA")
```
A map of the United States illustrating coffee price groups. States are colored in varying shades of brown to indicate whether their coffee prices are above or below the mean price. A dark grey outline defines each state, and a legend indicates the coffee price categories. We can see that "Below Mean" is more in the middle of the US. The "Above Mean" is on the west coast and on the upper east coast.

```{r}
library(dplyr)
library(tidyverse)
library(glue)
library(leaflet)
# Load spatial data for US states
library(sf)  # Load sf for spatial data
states <- read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson")  
class(states)  # Check the class of the states object

# Merge coffee data with state map data using spatial data
merged_data2 <- states %>%
  mutate(name = str_to_lower(name)) %>%
  left_join(coffee_data, by = c("name" = "State"))

# Create bins for coffee prices
bins <- c(0, 2.5, 3.0, 3.5, 5.0, Inf)
pal <- colorBin("YlOrRd", domain = merged_data$Price_of_Coffee, bins = bins)

# Create labels for hovering
merged_data2 <- merged_data2 %>%
  mutate(label = glue("{name}: ${Price_of_Coffee}"))

# Create the leaflet map
leaflet(data = merged_data2) %>%
  setView(-96, 37.8, 4) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~pal(Price_of_Coffee),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = ~label,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "3px 8px"),
      textsize = "15px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~Price_of_Coffee,
    opacity = 0.7,
    title = "Coffee Price ($)",
    position = "bottomright"
  )
```
An interactive map of the United States displaying coffee prices by state. Each state is filled with a color gradient representing coffee prices, with hover labels showing the state name and specific coffee price. A color legend in the bottom right corner indicates the price ranges, enhancing the visualization of coffee prices across the country. We can see the exact price for each state.